---
title: "Schnëssen vowels"
author: "Peter Gilles"
date: "21 4 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# certain requeries do not run with version 2.2.0
#require(devtools)
#install_version("emuR", version = "2.1.0", repos = "http://cran.us.r-project.org")
library(emuR)
library(tidyverse)
library(tidymodels)
library(joeyr)
library(tictoc)
library(phonR)
```

## References
[https://guilhermegarcia.github.io/vowels.html]
[https://marissabarlaz.github.io/portfolio/vowelnormalization/#speaker-intrinsicvowel-extrinsicformant-intrinsic]
[https://www.york.ac.uk/language/ypl/ypl2/11/YPL2-11-01-Flynn.pdf]
[https://rucforsk.ruc.dk/ws/portalfiles/portal/33209715/Website_FabriciusKendallWatt_NormalizationPlotting.pdf]
[https://drammock.github.io/phonR/]
[https://joeystanley.com/blog/making-vowel-plots-in-r-part-1]
[https://github.com/drammock/diphthong-plotting-tutorial/blob/master/diphthong-plotting-R.ipynb]
[https://www-users.york.ac.uk/~gb1055/sociophonetics_workshop/index.html]
[https://www.bas.uni-muenchen.de/~jmh/lehre/Rdf/EMU-SDMS/lesson8/08_Analysis-of-Formant-data2.html]
(Vowel Space)[https://rstudio-pubs-static.s3.amazonaws.com/348434_89d7208104214419b406a6f15d0d8bf3.html]
[https://icphs2019.org/icphs2019-fullpapers/pdf/full-paper_835.pdf]
[https://www.languagebits.com/files/euclidean-distance-diphthongs-in-r-code.html]
Renwick/Stanley 
[https://joeystanley.com/downloads/190103-lsa2019.pdf]
[https://asa.scitation.org/doi/pdf/10.1121/2.0000582]
[https://scholar.google.com/citations?view_op=view_citation&hl=en&user=EM9mqqMAAAAJ&sortby=pubdate&citation_for_view=EM9mqqMAAAAJ:Bg7qf7VwUHIC]
[https://asa.scitation.org/doi/pdf/10.1121/1.4894063] - Vergleich von Merger-Messmethoden
https://nzilbb.github.io/Covariation_monophthongs_NZE/Covariation_monophthongs_analysis.html

## Load EMU DB

```{r}
db <- load_emuDB("/Users/peter.gilles/Documents/_Daten/emuDBs/schnëssen_emuDB")
summary(db)

list_ssffTrackDefinitions(db)
list_levelDefinitions(db)
list_linkDefinitions(db)
```

## Add label groups
```{r}
# add short vowels label group
add_attrDefLabelGroup(db,
                      levelName = "MAU",
                      attributeDefinitionName = "MAU",
                      labelGroupName = "shortVowels",
                      labelGroupValues = c("i", "e", "ɛ", "æ", "ɑ", "ɔ", "o", "u", "ə"))

# add long vowels label group
add_attrDefLabelGroup(db,
                      levelName = "MAU",
                      attributeDefinitionName = "MAU",
                      labelGroupName = "longVowels",
                      labelGroupValues = c("iː", "eː", "ɛː", "aː", "oː", "uː"))

# add diphthongs label group
add_attrDefLabelGroup(db,
                      levelName = "MAU",
                      attributeDefinitionName = "MAU",
                      labelGroupName = "diphthongs",
                      labelGroupValues = c("iə", "əʊ", "uə", "ɜɪ", "ɑʊ", "æːɪ", "ɑɪ", "æːʊ"))

# add_labelGroup(db,
#                name = "monophthongs",
#                values = c("iː", "eː", "ɛː", "aː", "oː", "uː", "i", "e", "æ", "ɑ", "o", "ɔ", "u", "ə", "ˈə", "ɐ"))
# 
# add_labelGroup(db,
#                name = "diphthongs",
#                values = c("iə", "əʊ", "uə", "ɜɪ", "ɑʊ", "æːɪ", "ɑɪ", "æːʊ"))
# 
# add_labelGroup(db,
#                name = "shortVowels",
#                values = c("i", "e", "æ", "ɑ", "ɔ", "o", "u", "ə", "ˈə", "ɐ"))
#
# 
# add_labelGroup(db,
#                name = "longVowels",
#                values = c("iː", "eː", "ɛː", "aː", "oː", "uː"))
# 
# 
# list current label groups
list_attrDefLabelGroups(db,
                        levelName = "MAU",
                        attributeDefinitionName = "MAU")
# #
remove_attrDefLabelGroup(db,
                          levelName = "MAU",
                          attributeDefinitionName = "MAU",
                         labelGroupName = "monophthongs")
# # 
# # list label groups
list_labelGroups(db)
# # 
# # remove
# remove_labelGroup(db,
#                   name ="monophthongs")
```

## Load Sozialdaten

```{r}
sozialdaten <- rio::import("sozialdaten.csv") %>%
  rename(id = ID) %>%
  #filter(!Muttersprache == "Neen") %>%
  filter(!Geschlecht == "Aner") %>%
  filter(!Dialektgebiet == "") %>%
  select(id, Alter, Geschlecht, Dialektgebiet) %>%
  mutate(id  = as.character(id)) %>%
  mutate(Alter = as.factor(Alter)) %>%
  mutate(Geschlecht = as.factor(Geschlecht)) %>%
  mutate(Dialektgebiet = as.factor(Dialektgebiet)) %>%
  # define max frequency parameters for formant extraction for male and female
  mutate(ceiling = ifelse(Geschlecht == "Weiblech", 5500.0, 5000.0)) 
# %>%
#  mutate(Alter = recode(Alter, `65+` = '55+')) %>%
#  mutate(Alter = recode(Alter, `55 bis 64` = '55+'))
```


## Create formant tracks and pitch tracks by Praat (if necessary)

```{r}
# insert formant & pitch tracks, coming from praat
# see https://ips-lmu.github.io/The-EMU-SDMS-Manual/app-chap-wrassp.html#to-formant-burg-to-ssff-files
#source("https://raw.githubusercontent.com/IPS-LMU/The-EMU-SDMS-Manual/master/R/praat2AsspDataObj.R")
source("praat2AsspDataObj.R")

# list all .wav files in the emuDB
paths2wavFiles = list.files("/Users/peter.gilles/Documents/_Daten/emuDBs/99_bass_du_ëmmer_emuDB", pattern = "*.wav$", recursive = TRUE, full.names = TRUE)

# loop through files
files <- as.data.frame(paths2wavFiles) %>%
  mutate(id = word(word(paths2wavFiles, 9, sep ="/"), 3, sep = "_"))

filenames.meta <<- right_join(sozialdaten, files)

## formants
print("Now creating formant tracks...")

# function to submit more parameters to Praat than foreseen in emuR
formants <- function(file, ceiling) {
  # when 'ceiling' is eidel, dann default-Wert nehmen
  if (is.na(ceiling)) {
    ceiling <- 5500.0
  }
  ado = praat2AsspDataObj(file, object = "formant", ceiling = ceiling)
  newPath = paste0(tools::file_path_sans_ext(file), '.praatFms')
  #print(paste0(file, ' -> ', newPath)) # uncomment for simple log
  wrassp::write.AsspDataObj(ado, file = newPath)
}

# run function with two arguments with map2
map2(filenames.meta$paths2wavFiles, filenames.meta$ceiling, formants)

# # old from emuR doc
# for(fp in filenames.meta){
#   head(print(fp))
#   ado = praat2AsspDataObj(fp$paths2wavFiles, object = "formant", ceiling = fp$ceiling)
#   newPath = paste0(tools::file_path_sans_ext(fp), '.praatFms')
#   #print(paste0(fp, ' -> ', newPath)) # uncomment for simple log
#   wrassp::write.AsspDataObj(ado, file = newPath)
# }

## pitch
# print("Now creating pitch tracks...")
# for(fp in paths2wavFiles){
#   ado = praat2AsspDataObj(fp,
#                           object = "pitch",
#                           time_step = 0.00625,
#                           pitch_floor = 60,
#                           smooth = TRUE)
#   newPath = paste0(tools::file_path_sans_ext(fp), '.praatF0')
#   # print(paste0(fp, ' -> ', newPath)) # uncomment for simple log
#   wrassp::write.AsspDataObj(ado, file = newPath)
# }

# add SSFF track definition
add_ssffTrackDefinition(db,
                        name = "praatFms",
                        columnName = "fm",
                        fileExtension = "praatFms")

# add_ssffTrackDefinition(db,
#                         name = "praatF0",
#                         columnName = "f0",
#                         fileExtension = "praatF0")
```


## Get all KAN items

```{r}
# all words: KAN & ORT
sl <- query(db, query = "[ KAN =~ .*]", verbose=T) 
# attach a new column with the ORT labels using requery
sl2 <- sl %>%
  mutate(ORT = requery_hier(db, seglist = sl, level = "ORT")$labels)

# df with types for labels (=KAN) and ORT with frequency
all_KAN <- sl2 %>%
  count(labels, ORT, sort = T)
all_KAN
```

## Get all MAU items

```{r}
# get frequencies of all segments: MAU
sl <- query(db, "[ MAU =~ .*]")
sl2 <- sl %>%
  mutate(KAN = requery_hier(db, seglist = sl, level = "KAN")$labels) %>%
  mutate(ORT= requery_hier(db, seglist = sl, level = "ORT")$labels)
all_MAU <- sl2 %>%
  dplyr::select(labels, KAN, ORT) %>%
  count(labels, KAN, ORT, sort = T)
all_MAU
```

## Function to find and change the label for stressed vowels
The vowel will the have a leading ', e.g. 'ə

```{r}
# # funktioniert nicht richtig!
# # function to find stressed vowels
# get_stressed_vowels <- function(vowel) {
#   # al KAN segments containing stressed vowels
#   query_stressed <- paste0("[ KAN =~ '.*ˈ .*", vowel, " .*' ]")
#   sl_kan <- query(db, query = query_stressed)
#   sl_kan <- sl_kan %>%
#     distinct(labels)
#   # create a regex string from the vector, i.e. word|word|...
#   sl_kan <- as.vector(sl_kan$labels) %>%
#     paste(collapse = "|")
#   
#   # all MAU vowels only for these KAN words
#   query_stressed <- paste0("[ #MAU == '", vowel, "' ^ KAN =~ '(", sl_kan, ")' ]")
#   sl_mau <- query(db, query = query_stressed) 
#   # change labels for stressed vowel, add ˈ
#   toChange <- sl_mau %>% 
#     select(session, bundle, level, start_item_seq_idx, attribute, labels) %>% 
#     mutate(labels = paste0("ˈ", vowel))
#   # update db
#   update_itemsInLevel(db, toChange, verbose = TRUE)
#   # attach KAN and ORT columns
#   #sl_mau %>%
#   #  mutate(KAN = requery_hier(db, seglist = sl_mau, level = "KAN", collapse = TRUE)$labels) %>%
#   #  mutate(ORT = requery_hier(db, seglist = sl_mau, level = "ORT", collapse = TRUE)$labels)
# }
# 
# # has to be done one vowel-by-vowel basis
# stressed_vowels <- get_stressed_vowels(vowel = "ə")
```

## Getting the vowel data
## Prepare a df with all vowels and do time normalisation
```{r include=FALSE}
# construct search string
#query <- "[MAU == ˈə ^ ORT =~ '(dënschdeg|dënschden|dënsdeg|dënsden|gëschter|schëlder|schëlter|ëmmer|gëtt|schwëster|zëmmer|mëllech|nët|geschwënn)' ]"

#query <- "ORT == 'kanner'"
#query <- "MAU =~ '(i|iː|e|eː|ɛː|æ|aː|ɑ)'"
#query <- "MAU =~ '^(æ|aː)$'"
# query <- "MAU =~ '^(i|e|æ|ɑ|ɔ|o|u|ə)$'"
# query <- "MAU =~ '^(iː|eː|ɛː|aː|oː|uː)$'"
# query <- "MAU =~ '^(iə|əʊ|uə|ɜɪ|ɑʊ|æːɪ|ɑɪ|æːʊ)$'"
# query <- "MAU == shortVowels" 
# query <- "MAU == longVowels" 
# query <- "MAU == diphthongs" 

# function
get_segments <- function(query, max_dur, min_dur) {
  
vowel <- query(db, query = paste0("MAU == ", query)) 

stopwords <- rio::import("stopwords.txt")
vowel2 <- vowel %>%
 mutate(next_sound = requery_seq(db, vowel, offset = 1, ignoreOutOfBounds = TRUE)$labels) %>%
 mutate(previous_sound = requery_seq(db, vowel, offset = -1, ignoreOutOfBounds = TRUE)$labels)

vowel3 <- vowel2 %>%
  mutate(ORT = requery_hier(db, vowel, level = "ORT")$labels) %>%
  # column for stopwords  
  mutate(stopword = if_else(ORT %in% stopwords$stopword, "stopword", "content word")) %>%
  # filter out stopwords
  filter(stopword != "stopword") %>%
  # filter too short and too long vowels
  filter(end - start <= max_dur & end - start >= min_dur) %>%
  # filter problematic words with no formant values
  filter(ORT != "europa") %>%
  filter(ORT != "ugedo") %>%
  filter(ORT != "gema") %>%
  filter(ORT != "nooste") %>%
  filter(ORT != "nootste") %>%
  filter(ORT != "nächste") %>%
  filter(ORT != "gefalle")

# get formant tracks
td_vowels = get_trackdata(db, vowel3,
                          ssffTrackName = "praatFms", 
                          verbose = TRUE)

td_vowels_norm = normalize_length(td_vowels)

# save intermediate dfs / segment lists
saveRDS(td_vowels_norm, file = paste0("td_vowels_norm_", query, ".RDS"))

td_vowels_norm
}

# function calls for short, long vowels and diphthongs 
td_vowels_norm_shortVowels <- get_segments(query = "shortVowels", max_dur = 200, min_dur = 40)
td_vowels_norm_longVowels <- get_segments(query = "longVowels", max_dur = 350, min_dur = 80)
td_vowels_norm_diphthongs <- get_segments(query = "diphthongs", max_dur = 350, min_dur = 80)

# # read them
# td_vowels_shortVowels <- readRDS("td_vowels_shortVowels.RDS")
# td_vowels_longVowels <- readRDS("td_vowels_longVowels.RDS")
# td_vowels_diphthongs <- readRDS("td_vowels_diphthongs.RDS")
# 
# # normalise the length
# td_vowels_norm_shortVowels = normalize_length(td_vowels_shortVowels)
# td_vowels_norm_longVowels = normalize_length(td_vowels_longVowels)
# td_vowels_norm_diphthongs = normalize_length(td_vowels_diphthongs)
# # Achtung: wenn colNames verwendet wird, dann werden zusätzliche Metadatenspalten einfach gelöscht
# # die Warnung: "Found additional column ..." kann ignoriert werden
# # td_vowels_norm = normalize_length(td_vowels, colNames = c("T1", "T2"))
# 
# # save intermediate dfs / segment lists
# saveRDS(td_vowels_norm_shortVowels, "td_vowels_norm_shortVowels.RDS")
# saveRDS(td_vowels_norm_longVowels, "td_vowels_norm_longVowels.RDS")
# saveRDS(td_vowels_norm_diphthongs, "td_vowels_norm_diphthongs.RDS")

# bind and save segmentlist of whole database as df
td_vowels_norm <- bind_rows(td_vowels_norm_longVowels, td_vowels_norm_shortVowels, td_vowels_norm_diphthongs)

# save normalised df with all vowels
saveRDS(td_vowels_norm, "td_vowels_norm.RDS")

# check what is inside
td_vowels_norm %>%
  # to get only one row per segment
  filter(times_norm == 0.0) %>%
  group_by(labels, ORT) %>%
  count()
```

## Get vowel data for monophthong analysis
```{r}
# read entire  normalised df
#td_vowels_norm <- readRDS("td_vowels_norm.RDS")

# function to get formants for specific point in time ('times_norm')
get_formants <- function(data, time, segments){
  
  # tibble with words to be expluded
  source(file.path(paste0(getwd()), "excluded_words.R"))
  
  data <- data %>% 
    # exclude words from list
    filter(!ORT %in% excluded_words$excluded) %>%
    # filter labels
    dplyr::filter(str_detect(labels, segments)) %>%
    # select a point 
    dplyr::filter(times_norm >= time & times_norm <= time + 0.04) %>%
    # filter implausible/wrong formant values
    #filter(T1 >= 200 & T1 <= 1000) %>%
    #filter(T2 <= 3000) %>%
    # duration
    mutate(duration = round(end - start))
  }

# ɔ excluded
formants_mid_monophthongs <- get_formants(data = td_vowels_norm, 
                                          time = 0.3,
                                          segments = "^æ$|aː|ɑ$|iː$|i$|uː|u$|o|eː$|ɛː|ɛ$|e|o|ə$")
formants_onset_diphthongs <- get_formants(data = td_vowels_norm, 
                                          time = 0.1,
                                          segments = "æːɪ|æːʊ|ɑɪ|ɑʊ|ɜɪ|əʊ|iə|uə")

# filtering implausible formant values
# reduces the df by ca. 15 %
formants_all <- bind_rows(formants_mid_monophthongs, formants_onset_diphthongs) %>%
    # create id for speaker
    mutate(id = word(bundle, 3, sep ="_")) %>%
  dplyr::filter(labels == "iː" & T1 <= 500 & T2 <= 3000 | 
                labels == "i" & T1 <= 500 & T2 <= 3000 | 
                labels == "eː" & T1 <= 600 & T2 <= 2900 | 
                labels == "e" & T1 <= 700 & T2 <= 2800 | 
                labels == "ə" & T1 <= 800 & T2 <= 2000 | 
                labels == "ɜɪ" & T1 <= 700 & T2 <= 2500 |
                labels == "ɑ" & T1 <= 900 & T2 <= 1600 | 
                labels == "ɑɪ" & T1 <= 900 & T2 <= 1700 | 
                labels == "ɑʊ" & T1 <= 900 & T2 <= 1400 | 
                labels == "aː" & T1 <= 1200 & T2 <= 2000 | 
                labels == "ɛː" & T1 <= 800 & T2 <= 2300 | 
                labels == "ɛ" & T1 <= 800 & T2 <= 2300 | 
                labels == "æ" & T1 <= 1000 & T2 <= 2100 | 
                labels == "æːɪ" & T1 <= 1000 & T2 <= 2000 | 
                labels == "æːʊ" & T1 <= 1000 & T2 <= 1800 | 
                labels == "o" & T1 <= 700 & T2 <= 1300 | 
                labels == "ɔ" & T1 <= 800 & T2 <= 1300 | 
                labels == "oː" & T1 <= 700 & T2 <= 1300 | 
                labels == "u" & T1 <= 500 & T2 <= 1100 | 
                labels == "əʊ" & T1 <= 700 & T2 <= 1600 | 
                labels == "iə" & T1 <= 500 & T2 <= 3000 | 
                labels == "uə" & T1 <= 500 & T2 <= 1100)

#Lobanov 2.0 - calculate means per vowel and per speaker
summary_vowels_all <- formants_all %>%
  group_by(id, labels) %>%
  dplyr::summarise(mean_F1 = mean(T1),
            mean_F2 = mean(T2),
            sd_F1 = sd(T1),
            sd_F2 = sd(T2),
            token_count_vowel = n())

#get the mean_of_means and sd_of_means from the the speaker_summaries, this will give each speaker a mean caculated from the means across all vowels, as well as the standard deviation of the means
summary_mean_of_means <- summary_vowels_all %>%
  group_by(id) %>%
  dplyr::summarise(mean_of_means_F1 = mean(mean_F1),
            mean_of_means_F2 = mean(mean_F2),
            sd_of_means_F1 = sd(mean_F1),
            sd_of_means_F2 = sd(mean_F2)
            )

# normalisation
formants_all <- formants_all %>%
    # create id for speaker
    mutate(id = word(bundle, 3, sep ="_")) %>%
    # join Sozialdaten
    inner_join(sozialdaten) %>%
    # remove outliers
    group_by(Alter, Geschlecht, labels) %>%
    filter(!find_outliers(T1, T2, keep = 0.95)) %>%
    ungroup() %>%
    # normalise with Lobanov 2.0 (https://nzilbb.github.io/Covariation_monophthongs_NZE/Covariation_monophthongs_analysis.html#Normalisation)
    #add in the data
    left_join(., summary_mean_of_means) %>%
    left_join(., summary_vowels_all[, c("id", "labels", "token_count_vowel")]) %>%
    #normalise with Lobanov 2.0
    mutate(F1_lob2 = (T1 - mean_of_means_F1)/sd_of_means_F1,
           F2_lob2 = (T2 - mean_of_means_F2)/sd_of_means_F2) %>%
    #remove the variables that are not required
    dplyr::select(-(mean_of_means_F1:sd_of_means_F2)) %>%
    group_by(id) %>%
    # normalise with Lobanov per id, i.e. on the basis of the individual speaker
    mutate(F1_lob = normLobanov(T1), F2_lob = normLobanov(T2), F3_lob = normLobanov(T3)) %>%
    # normalise with deltaF (Johnson 2020)
    norm_deltaF(T1, T2, T3, T4) %>%
    rename(F1_deltaF = T1_deltaF, F2_deltaF = T2_deltaF, F3_deltaF = T3_deltaF, F4_deltaF = T4_deltaF) %>%
    ungroup() %>%
    # keep only speakers with more than 40 (?) segments  
    group_by(id) %>%
    filter(n() >= 40) %>%
    ungroup()
    
#remove the previous summary data frames
rm(summary_vowels_all, summary_mean_of_means)

# save formants_all
saveRDS(formants_all, "formants_all.RDS")

# which words per vowel
formants_all %>% 
  filter(labels == "ɑ") %>% 
  group_by(ORT, labels) %>% 
  count()

formants_speaker_median <- formants_all %>%
    group_by(labels, id, Alter, Geschlecht, Dialektgebiet) %>%
    # calculate median for T1, T2, T3 and duration
    summarise_at(vars(T1:T3, F1_lob, F2_lob, F3_lob, F1_deltaF, F2_deltaF, F3_deltaF, duration), median) %>%
    ungroup()
```


## Plot individual vowel system
```{r}
plot_individual_chart <- function(data) {
  # calculate centroid 
  temp <<- data %>%
  group_by(labels)
  
  ggplot(temp) +
    aes(x = T2, y = T1, label = labels, col = ORT) +
    #aes(x = F2_lob, y = F1_lob, label = labels, col = labels) +
    #geom_text() +
    geom_point() +
    scale_y_reverse() +
    scale_x_reverse() +
    labs(x = "F2", y = "F1") 
}

formants_all %>%
  #filter(ORT == "neisten") %>%
  filter(labels == "æ") %>%
  #filter(id == 200) %>%
  plot_individual_chart()
```



## Plot per age, six plots per age group
```{r}
library(ggpointdensity)
library(viridis)

data <- formants_all %>%
  # exclude certain vowels
  #filter(labels %in% c("ɑ"))
  #filter(labels %in% c("ə", "i", "aː", "u"))
  #filter(labels %in% c("æːɪ", "ɜɪ", "aː", "eː", "ɛː", "e", "i", "u"))
  filter(labels %in% c("æːɪ", "ɑɪ", "æːʊ", "ɑʊ", "aː", "ɑ", "æ", "ɜɪ", "əʊ", "i", "u", "eː", "ɛː", "oː", "o", "ɔ"))

plot_sozialdaten <- function(data, social) {
  # calculate centroid 
  temp <- data %>%
  group_by(Alter, Geschlecht, labels) %>%
  mutate(T1_centroid = median(T1), T2_centroid = median(T2)) %>%
  mutate(F1_lob_centroid = median(F1_lob), F2_lob_centroid = median(F2_lob)) %>%
  mutate(F1_deltaF_centroid = median (F1_deltaF), F2_deltaF_centroid = median(F2_deltaF))
  
  ggplot(temp) +
    #aes(x = T2, y = T1, label = labels, col = labels) +
    aes(x = F2_lob2, y = F1_lob2, label = labels, col = labels) +
    #aes(x = F2_deltaF, y = F1_deltaF, label = labels, col = labels) +
    #geom_text() +
    #geom_density2d() +
    # not really working
    #geom_pointdensity(aes(x = F2_deltaF, y = F1_deltaF, label = labels, col = labels)) +
    #scale_color_viridis() +
    # stat_density2d(geom = "polygon",
    #               aes(alpha = ..level.., fill = labels),
    #               bins = 4) +
    # add centroids per vowel
    #geom_text(aes(x =T2_centroid, y = T1_centroid, label = labels, col = labels)) +
    geom_text(aes(x =F2_lob_centroid, y = F1_lob_centroid, label = labels, col = labels),               size = 5) +
    #geom_text() +
    # centroid of entire vowel system
    #geom_text(aes(x =F2_lob_centroid_speaker, y = F1_lob_centroid_speaker, label = "C"),               size = 5) +
    #stat_ellipse() +
    scale_y_reverse() +
    scale_x_reverse() +
    labs(x = "F2", y = "F1") +
    facet_grid({{social}} ~ Alter)
}

# plot centroids per vowel
plot_sozialdaten(data, data$Geschlecht)

# how many vowels per group?
data %>%
  group_by(Geschlecht, Alter) %>%
  count()

# how many vowels?
data %>%
  group_by(labels) %>%
  count(sort = T)

# # playing with PhonR illustrations
# plots <- function (data, Geschlecht, Alter) {
#   plotVowels(data$F1_lob, data$F2_lob, data$labels, group = FALSE, plot.tokens = FALSE, plot.medians = TRUE,
#     pch.medians = data$labels, cex.medians = 2, var.col.by = data$labels, var.sty.by = FALSE, ellipse.fill = TRUE,
#     poly.line = TRUE, poly.order = c("ɜɪ", "æːɪ", "æ", "aː", "ɑ", "ɑɪ"), pretty = TRUE,
#     sub = paste(Geschlecht, Alter))
# }
# 
# plots <- data %>%
#   #dplyr::arrange(Geschlecht) %>%
#   select(Geschlecht, Alter, labels, F1_lob, F2_lob) %>%
#   group_by(Geschlecht, Alter) %>%
#   nest() %>%
#   mutate(plots = purrr::pmap(list(Geschlecht = Geschlecht,
#                                    Alter = Alter,
#                                    data = data),
#                                    plots))
```

## Plot per age in one plot with arrows
```{r}
data <- formants_all %>%
  # exclude certain vowels
  #filter(labels %in% c("ɑ"))
  #filter(labels %in% c("ə", "i", "aː", "u"))
  #filter(labels %in% c("æːɪ", "ɜɪ", "aː", "eː", "ɛː", "e", "i", "u"))
  filter(labels %in% c("æːɪ", "ɑɪ", "æːʊ", "ɑʊ", "aː", "ɑ", "æ", "ɜɪ", "əʊ", "i", "u", "eː", "ɛː", "oː", "o", "ɔ")) %>%
  mutate(Alter = recode(Alter, `≤ 24` = 2003, `25 bis 34` = 1993, `35 bis 44` = 1983, `45 bis 54` = 1973, `55 bis 64` = 1963, `65+` = 1953, `55+` = 1963)) %>%
    # calculate centroid 
    group_by(labels, Geschlecht, Alter) %>%
    summarize(T1 = median(T1), T2 = median(T2), F1_lob = median(F1_lob), F2_lob = median(F2_lob), F1_lob2 = median(F1_lob2), F2_lob2 = median(F2_lob2), F1_deltaF = median (F1_deltaF), F2_deltaF = median(F2_deltaF))

sound_change_labels1 <- data %>%
  group_by(Geschlecht, labels) %>%
  filter(Alter == min(Alter))

#make data frame to plot end point, this will give the arrow at the end of the paths based on the largest year of birth coordinates
sound_change_labels2 <- data %>%
  group_by(Geschlecht, labels) %>%
  top_n(wt = Alter, n = 2)

# plot tracks for changes over age
ggplot(data, aes(x =F2_lob, y = F1_lob, group = labels, alpha = Alter, col = labels)) +
  geom_path(size = 1, show.legend = FALSE) +
  geom_path(data = sound_change_labels2, aes(x = F2_lob, y = F1_lob, colour = labels, group = labels), arrow = arrow(ends = "last", type = "closed", length = unit(0.2, "cm")), inherit.aes = FALSE, show.legend = FALSE) +
 geom_text(data = sound_change_labels1, aes(x = F2_lob, y = F1_lob, colour = labels, group = labels, label = labels), inherit.aes = FALSE, show.legend = FALSE) +
  scale_x_reverse() +
  scale_y_reverse() +
  facet_wrap(~ Geschlecht)

```


## Sound change animation
```{r}
library(gganimate)
sound_change_plot_animation <- data %>%
  #set general aesthetics
  ggplot(aes(x = F2_lob, y = F1_lob, colour = labels, group = labels, label = labels)) +
  geom_text(aes(fontface = 2), size = 8, show.legend = FALSE) +
  # geom_point() +
  geom_path() +
  #label the axes
  xlab("F2 (normalised)") +
  ylab("F1 (normalised)") +
  #reverse the axes to follow conventional vowel plotting
  scale_x_reverse() +
  scale_y_reverse() +
  #scale_x_reverse(limits = c(2,-2), position = "top") +
  #scale_y_reverse(limits = c(2,-2), position = "right") +
  #set the colours
  scale_color_manual(values = c("#9590FF", "#D89000", "#A3A500", "#39B600", "#00BF7D",
                                 "#00BFC4", "#00B0F6", "#F8766D", "#E76BF3", "#FF62BC",
                                "#a6cee3","#1f78b4","#b2df8a","#33a02c", "#33a02c")) +
  #add a title
  labs(caption = 'Year of birth (interpolated): {round(frame_along, 0)}') +
  #set the theme
  theme_bw() +
  #make text more visible
  theme(axis.title = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(size = 16, face = "bold"),
        axis.text.y = element_text(size = 16, face = "bold", angle = 270),
        axis.ticks = element_blank(),
        plot.caption = element_text(size = 20, hjust = 0),
        legend.position = "none") +
  #set the variable for the animation transition i.e. the time dimension
  transition_reveal(Alter) +
  #add in a trail to see the path
  shadow_trail(max_frames = 100, alpha = 0.1) +
  ease_aes('linear') +
  facet_wrap(~ Geschlecht)

sound_change_plot_animation <- animate(sound_change_plot_animation, nframes = 200, fps = 5, rewind = FALSE, start_pause = 10, end_pause = 10, duration = 20, height = 600, width =1000)

anim_save(sound_change_plot_animation, filename = "lux_vowels_change_animation.gif")

sound_change_plot_animation
```


## Plot Formants x Alter
```{r}
data <- formants_all %>%
         filter(labels == "ɑ")
ggplot(data,
       aes(x= Alter, y = F1_lob2)) +
  geom_boxplot() +
  facet_wrap(~ Geschlecht)

library(nlme)
m0 <- lme(F1_lob2 ~1, random = ~ 1|id/ORT, data = data, method = "ML")
m1 <- update(m0, .~. + Alter)
m2 <- update(m1, .~. + Geschlecht)
m3 <- update(m1, .~. + Alter:Geschlecht)

anova(m0, m1, m2, m3)
summary(m2)

regression <- lm(formula = F1_lob2 ~ Alter, data = data)
summary(regression)
```

## GAMMs

https://nzilbb.github.io/Covariation_monophthongs_NZE/Covariation_monophthongs_analysis.html#GAMM_modelling
```{r}
vowels_all <- formants_all %>%
  mutate(participant_year_of_birth = recode(Alter, `≤ 24` = 2003, `25 bis 34` = 1993, `35 bis 44` = 1983, `45 bis 54` = 1973, `55 bis 64` = 1963, `65+` = 1953, `55+` = 1963)) %>%
  rename(Vowel = labels, F1_lobanov_2.0 = F1_lob2, F2_lobanov_2.0 = F2_lob2,
         Speaker =id,
         Word = ORT,
         Gender = Geschlecht) %>%
  mutate(Gender = recode(Gender, `Weiblech` = "F")) %>%
  mutate(Gender = recode(Gender, `Männlech` = "M")) %>%
  mutate(participant_year_of_birth = as.integer(participant_year_of_birth)) %>%
  mutate(Speaker = as.factor(Speaker)) %>%
  mutate(Word = as.factor(Word)) %>%
  mutate(Vowel = as.factor(Vowel))

#update the Gender variable to allow for conrast coding
vowels_all$Gender <- as.ordered(vowels_all$Gender)
contrasts(vowels_all$Gender) <- "contr.treatment"

vowels_all <- vowels_all %>%
  arrange(as.character(Speaker))

vowels_all <- vowels_all %>%
  select(Speaker, Gender, participant_year_of_birth, Word, Vowel, F1_lobanov_2.0, F2_lobanov_2.0) %>%
  filter(Vowel %in% c("aː", "æ", "ɑ")) 

# speakers having ALL required vowels
vowels_all <- vowels_all %>% 
  group_by(Speaker) %>% 
  mutate(tokens_per_speaker = n()) %>%
  ungroup() %>%
  group_by(Speaker, Vowel) %>% 
  mutate(vowels_per_speaker = n()) %>%
  filter(tokens_per_speaker >= 20) %>%
  ungroup()
```

```{r}
library(mgcv)
library(itsadug)
#create a data frame to store the intercepts from the models, this will initially contain just the speaker names
gam_intercepts.tmp <- vowels_all %>%
  dplyr::select(Speaker) %>%
  distinct()

#loop through the vowels
cat(paste0("Start time:\n", format(Sys.time(), "%d %B %Y, %r\n")))

#vowels_all <- vowels_nz
for (i in levels(factor(vowels_all$Vowel))) {
  
  #F1 modelling
  
  #run the mixed-effects model on the vowel, i.e. if i = FLEECE this will model F1 for FLEECE
  gam.F1 <- bam(F1_lobanov_2.0 ~
                  s(participant_year_of_birth, k=5, bs="cr", by=Gender) +
                  s(participant_year_of_birth, k=5, bs="cr") +
                  Gender +
                  #s(Speech_rate) +
                  s(Speaker, bs="re") +
                  s(Word, bs="re"),
                data=vowels_all %>% filter(Vowel == i),
                discrete=T, nthreads=2)
  
  #extract the speaker intercepts from the model and store them in a temporary data frame
  gam.F1.intercepts.tmp <- as.data.frame(get_random(gam.F1)$`s(Speaker)`)
  
  #assign the model to an object
  assign(paste0("gam_F1_", i), gam.F1)
  
  #save the model summary
  saveRDS(gam.F1, file = paste0("gam_F1_", i, ".rds"))
  
  cat(paste0("F1_", i, ": ", format(Sys.time(), "%d %B %Y, %r"), " ✅\n")) #print the vowel the loop is up to for F1, as well as the start time for the model
  
  #F2 modelling
  
  #run the mixed-effects model on the vowel, i.e. if i = FLEECE this will model F2 for FLEECE
  gam.F2 <- bam(F2_lobanov_2.0 ~
                  s(participant_year_of_birth, k=5, bs="cr", by=Gender) +
                  s(participant_year_of_birth, k=5, bs="cr") +
                  Gender +
                  #s(Speech_rate) +
                  s(Speaker, bs="re") +
                  s(Word, bs="re"),
                data=vowels_all %>% filter(Vowel == i),
                discrete=T, nthreads=2)
  
  #extract the speaker intercepts again, storing them in a separate data frame
  gam.F2.intercepts.tmp <- as.data.frame(get_random(gam.F2)$`s(Speaker)`)
  
  #assign the model to an object
  assign(paste0("gam_F2_", i), gam.F2)
  
  #save the model summary
  saveRDS(gam.F2, file = paste0("gam_F2_", i, ".rds"))

  #rename the variables so it clear which one has F1/F2, i.e. this will give F1_FLEECE, F2_FLEECE
  names(gam.F1.intercepts.tmp) <- paste0("F1_", i)
  names(gam.F2.intercepts.tmp) <- paste0("F2_", i)
  
  #combine the intercepts for F1 and F2 and store them in the intercepts.tmp_stress data frame
  gam_intercepts.tmp <- cbind(gam_intercepts.tmp, gam.F1.intercepts.tmp, gam.F2.intercepts.tmp)
  
  cat(paste0("F2_", i, ": ", format(Sys.time(), "%d %B %Y, %r"), " ✅\n")) #print the vowel the loop is up to for F2 , as well as the start time for the model
}

#save the intercepts as a .csv file
write.csv(gam_intercepts.tmp, "gam_intercepts_tmp_new.csv", row.names = FALSE)
```

```{r}
vowels_to_plot <- c("ɑ", "aː", "æ")

pred_table <- function (Vowel) {
  mod_name <- paste0("gam_F1_", Vowel)
  return(cbind(Vowel, 
               plot_smooth(get(mod_name), view="participant_year_of_birth", 
                           rm.ranef=T, rug=F, n.grid=500)$fv)
  )
}

gamm_preds_to_plot <- lapply(vowels_to_plot, pred_table) %>%
  do.call(rbind, .) %>%
  mutate(participant_year_of_birth = as.integer(participant_year_of_birth))

saveRDS(gamm_preds_to_plot, "gamm_preds_to_plot.rds")
```

```{r}
#make a long version of the intercepts
gam_intercepts.tmp_long <- gam_intercepts.tmp %>%
  pivot_longer(F1_aː:F2_ɑ, names_to = "Vowel_formant", values_to = "Intercept") %>%
  mutate(Formant = substr(Vowel_formant, 1, 2),
         Vowel = substr(Vowel_formant, 4, max(nchar(Vowel_formant)))) %>%
  left_join(vowels_all %>%
    select(Speaker, participant_year_of_birth) %>% distinct()) %>%
  left_join(gamm_preds_to_plot %>% 
              mutate(Formant = "F1") %>% 
  select(participant_year_of_birth, Vowel, Formant, fit, ll, ul))

speakers1 <- gam_intercepts.tmp_long %>%
  filter(Vowel_formant %in% c("F1_aː", "F1_æ", "F1_ɑ")) %>%
  ungroup() %>%
  arrange(participant_year_of_birth) %>%
  filter(Speaker %in% c("1000", "1630", "421", "1004")) %>%
  mutate(letters = c(rep("A", 3), rep("B", 3), rep("C", 3), rep("D", 3)),
         speakers1 = paste0(letters, " (", round(Intercept, 2), ")"))

F1_speaker_intercepts <- gam_intercepts.tmp_long %>%
  filter(Vowel_formant %in% c("F1_aː", "F1_æ", "F1_ɑ"),
         Formant == "F1") %>% #choose the vowels
  mutate(Vowel = factor(Vowel, levels = c("aː", "æ", "ɑ"))) %>% #order the vowels
  ggplot(aes(x = participant_year_of_birth, y = fit, colour = Vowel, fill = Vowel)) +
  geom_point(aes(x = participant_year_of_birth, y = fit + Intercept), size = 1, alpha = 0.2) +
  geom_line() +
  geom_ribbon(aes(ymin=ll, ymax=ul), colour = NA, alpha=0.2) +
  #geom_point(data = speakers1 %>% mutate(fit = ifelse(speakers1 %in% c("D (-0.36)"), fit - 0.05, fit)), aes(x = participant_year_of_birth, y = fit + Intercept), size = 4) +
  #geom_label(data = speakers1 %>% mutate(fit = ifelse(speakers1 %in% c("B (0.23)"), fit - 0.1, fit), fit = ifelse(speakers1 %in% c("D (-0.36)"), fit - 0.1, fit)), aes(x = participant_year_of_birth - 1, y = fit + Intercept, label = speakers1), fill = "white", alpha = 0.5, hjust = 1, show.legend = FALSE) +
  scale_y_reverse() +
  scale_colour_manual(values=c("#E69F00", "#56B4E9", "#009E73")) +
  scale_fill_manual(values=c("#E69F00", "#56B4E9", "#009E73")) +
  xlab("Participant year of birth") + #x axis title
  ylab("Normalised F1 (Lobanov 2.0)") + #y axis title
  theme_bw() + #general aesthetics
  theme(legend.position = c(0.8, 0.1), #legend position
        legend.direction = "horizontal",
        legend.background = element_rect(colour = "black"),
        axis.text = element_text(size = 14), #text size
        axis.title = element_text(face = "bold", size = 14), #axis text aesthetics
        legend.title = element_blank(), #legend title text size
        legend.text = element_text(size = 14)) #legend text size
 
F1_speaker_intercepts

ggsave(plot = F1_speaker_intercepts, filename = "F1_speaker_intercepts.png", width = 10, height = 5, dpi = 400)
```


## Visualisation of intercept correlations

```{r}
gam_intercepts.tmp %>%
  select(-1) %>%
  pivot_longer(cols = 1:6, names_to = "variable", values_to = "intercept") %>%
  ggplot(aes(x = intercept, y = ..density..)) +
  geom_histogram(bins = 500) +
  geom_density(outline.type = "upper", colour = "red", bw = "SJ") +
  facet_wrap(~variable) +
  theme_bw()
```

## Heat map of correlations

```{r}
cors <- function(df) { 
   # turn all three matrices (r, n, and P into a data frame)
   M <- Hmisc::rcorr(as.matrix(df))
   # return the three data frames in a list return(Mdf)
   Mdf <- map(M, ~data.frame(.x))
}

formatted_cors <- function(df){
 cors(df) %>%
 map(~rownames_to_column(.x, var="measure1")) %>%
 map(~pivot_longer(.x, -measure1, "measure2")) %>% 
 bind_rows(.id = "id") %>%
 pivot_wider(names_from = id, values_from = value) %>%
 mutate(sig_p = ifelse(P < .05, T, F), p_if_sig = ifelse(P <.05, P, NA), r_if_sig = ifelse(P <.05, r, NA))
}

formatted_cors(gam_intercepts.tmp[-1]) %>%
 ggplot(aes(x = measure1, y = measure2, fill = r, label=round(r_if_sig,2))) +
 geom_tile() +
 geom_text(aes(size = abs(r_if_sig)), show.legend = FALSE) +
 labs(x = NULL, y = NULL, fill = "Pearson's Correlation") +
 # map a red, white and blue color scale to correspond to -1:1 sequential gradient
 scale_fill_gradient2(mid="#FBFEF9",low="#0C6291",high="#A63446", limits=c(-1,1)) +
 theme_classic() +
 # remove excess space on x and y axes
 scale_x_discrete(expand=c(0,0)) +
 scale_y_discrete(expand=c(0,0)) +
 # change global font to roboto
 theme(axis.text = element_text(size = 20, face = "bold"),
       axis.text.x = element_text(angle = 90, hjust = 1),
       legend.position = "top",
       legend.key.width = unit(2, "cm"),
       legend.title = element_text(size = 20),
       legend.text = element_text(size = 20)) +
  guides(fill = guide_colourbar(title.position = "top"))
```

## Principal Components Analysis (PCA)

```{r}
library(factoextra)
#run the PCA on the intercepts data frame, note the intercepts are in columns 2:21 as column 1 is the speaker name
intercepts.pca <- princomp(gam_intercepts.tmp[, 2:ncol(gam_intercepts.tmp)],cor=TRUE)

#print a summary of the PCA, this will give the variance explained by each PC
summary(intercepts.pca)
```

```{r}
#view eigen values, this will visualise the variance explained by each PC
factoextra::fviz_eig(intercepts.pca, addlabels = TRUE)
```

```{r}
#create objects containing the loadings for each of the 3 main PCs
PC1_loadings <- intercepts.pca$loadings[,1]
PC2_loadings <- intercepts.pca$loadings[,2]
PC3_loadings <- intercepts.pca$loadings[,3]

#store these in a data frame for ease of plotting
PC1_loadings <- PC1_loadings %>%
  as.data.frame() %>%#convert to a data frame
  dplyr::rename(Loading = 1) %>% #rename the variable
  mutate(variable = row.names(.), #add variable to identify which intercepts are representing each row
         highlight = ifelse(Loading > 0.225 | Loading < -0.225 , "black", "gray"), #for plotting reasons, we want to highlight the variables that contribute the most to the PC, so if the loading is > |0.2| it will plotted in black, if not it will be gray)
         PC1_loadings_abs = abs(Loading), #make the loadings absolute so they are comparable when plotting
         direction = ifelse(Loading < 0, "red", "black"), #define which direction the loading sign was, i.e. red if negative, black if positive
         PC = "PC1",
         Vowel = substr(variable, 4, 10)) %>% #add variable to identify which PC the data is from
  arrange(PC1_loadings_abs) #order the variables based on absolute loading value

#repeat this process for PC2
PC2_loadings <- PC2_loadings %>%
  as.data.frame() %>%
  dplyr::rename(Loading = 1) %>%
  mutate(variable = row.names(.),
         highlight = ifelse(Loading > 0.225 | Loading < -0.225 , "black", "gray"),
         PC2_loadings_abs = abs(Loading),
         direction = ifelse(Loading < 0, "red", "black"),
         PC = "PC2",
         Vowel = substr(variable, 4, 10)) %>%
  arrange(PC2_loadings_abs)

#repeat this process for PC3
PC3_loadings <- PC3_loadings %>%
  as.data.frame() %>%
  dplyr::rename(Loading = 1) %>%
  mutate(variable = row.names(.),
         highlight = ifelse(Loading > 0.225 | Loading < -0.225 , "black", "gray"),
         PC3_loadings_abs = abs(Loading),
         direction = ifelse(Loading < 0, "red", "black"),
         PC = "PC3",
         Vowel = substr(variable, 4, 10)) %>%
  arrange(PC3_loadings_abs)
```

```{r}
PC_loadings_contrib <- intercepts.pca$loadings[,1:3] %>% #get the loadings for PC1, PC2 and PC3
  as.data.frame() %>%
  cbind(get_pca(intercepts.pca, "var")$contrib[,1:3]) %>% #get the contributions
  dplyr::rename(Loading.PC1 = Comp.1,
         Loading.PC2 = Comp.2,
         Loading.PC3 = Comp.3,
         Contribution.PC1 = Dim.1,
         Contribution.PC2 = Dim.2,
         Contribution.PC3 = Dim.3) %>%
  mutate(Variable = row.names(.),
         Vowel = substr(Variable, 4, nchar(Variable))) %>%
  dplyr::select(Variable, Vowel, Loading.PC1:Contribution.PC3) %>%
  pivot_longer(Loading.PC1:Loading.PC3, names_to = "PC", values_to = "Loading") %>%
  arrange(Vowel, Variable, PC) %>%
  mutate(direction = ifelse(Loading < 0, "red", "black")) #define direction of the loadings
```

### PC1

```{r}
PC1_contrib <- PC_loadings_contrib %>%
  select(-Contribution.PC2, -Contribution.PC3) %>%
  filter(PC == "Loading.PC1") %>%
  arrange(Contribution.PC1) %>%
  mutate(Loading_abs = abs(Loading),
         cumsum_PC1 = round(cumsum(Contribution.PC1), 4),
         highlight = ifelse(cumsum_PC1 < 50, "grey", "black"),
         highlight1 = ifelse(cumsum_PC1 < 50, 0.5, 1),
         direction1 = direction,
         direction_lab = ifelse(direction1=="red","–", "+"))
  
PC1_contrib_plot <- ggplot(PC1_contrib, aes(x=reorder(Variable, Contribution.PC1), y=Contribution.PC1)) + #have the variable name on the x axis and loading value on the y
  geom_text(aes(alpha = highlight, label=ifelse(PC1_contrib$direction=="red","–", "+")), colour = PC1_contrib$direction, size = 6, fontface="bold", show.legend = FALSE) + 
  xlab("") + #remove the x axis title
  ylab("PC1 contribution (%)") +
  geom_vline(xintercept = 16.5, color = "red", linetype = "dashed") + #add a red dashed line to identify the 0.2 cut off
  scale_alpha_manual(values=c(1, 0.3)) +
  theme_bw() + #set the aesthetics of the plot
  theme(axis.text.x = element_text(angle = 90, hjust = 1, colour = PC1_contrib$highlight, size = 14, face = "bold"),
        axis.text.y = element_text(size = 14, face = "bold"),
        axis.title = element_text(size = 14, face = "bold")) #modify the x axis variable names so they are rotated and highlighted

PC1_contrib_plot
```


## Visualisation by GAMM smooth

### Sound change by year
```{r}
#make vector containing all .rds filenames from model_summaries folder
model_summary_files = list.files(pattern="*.rds", path = "model_summaries")

#create a dummy data frame to store the values from plot_smooth
mod_pred_yob_values <- data.frame(plot_smooth(gam_F1_aː, view="participant_year_of_birth", rm.ranef=T)$fv) %>%
   mutate(model1 = "gam_F1_aː1") %>%
  filter(model1 != "gam_F1_aː1")

#load each of the files with for loop
for (i in model_summary_files) {
  
  #remove .rds from model name and then print the model name
  mod1 <- gsub(".rds", "", i)
  print(mod1)
  
  #generate a plot smooth of the model giving the participant year of birth smooth
  plot1 <- plot_smooth(get(mod1), view="participant_year_of_birth", rm.ranef = T, n.grid = 119, main = mod1)

  #extract the values from the plot and reformat for ggplot
  plot1 <- plot1$fv %>%
    mutate(Vowel = substr(mod1, 8, nchar(mod1)),
           Formant = substr(mod1, 5, 6))
  
  #combine the data to the previous rows, so that all models are in a single data frame
  mod_pred_yob_values <<- rbind(mod_pred_yob_values, plot1) %>%
    mutate(participant_year_of_birth = as.integer(participant_year_of_birth))
  
}

saveRDS(mod_pred_yob_values, "Models/mod_pred_yob_values.rds")
```

```{r}
#make a long version of the intercepts
gam_intercepts.tmp_long <- gam_intercepts.tmp %>%
  pivot_longer(F1_aː:F2_ɑ, names_to = "Vowel_formant", values_to = "Intercept") %>%
  mutate(Formant = substr(Vowel_formant, 1, 2),
         Vowel = substr(Vowel_formant, 4, max(nchar(Vowel_formant)))) %>%
  left_join(vowels_all %>%
    select(Speaker, participant_year_of_birth, Gender) %>% distinct()) %>%
  left_join(mod_pred_yob_values %>% select(participant_year_of_birth, Vowel, Formant, fit, ll, ul))

#plot the gam smooths predicting F1/F2 per vowel and add the speaker intercepts values
sound_change_plot_smooth <- mod_pred_yob_values %>%
  mutate(Formant = factor(Formant, levels = c("F2", "F1")))  %>%
  ggplot(aes(x = participant_year_of_birth, y = fit, colour = Formant, fill = Formant)) +
  geom_jitter(data = gam_intercepts.tmp_long %>% mutate(Formant = factor(Formant, levels = c("F2", "F1"))), aes(x = participant_year_of_birth, y = fit + Intercept), size = 0.25, alpha = 0.1) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = ll, ymax = ul), alpha = 0.2, colour = NA) +
  scale_x_continuous(breaks = c(1900, 1950)) +
  xlab("Year of birth") +
  ylab("Predicted model fit (Lobanov 2.0)") +
  facet_grid(Formant~Vowel) +
  theme_bw() +
  theme(legend.position = "none", axis.text = element_text(size = 12), axis.title = element_text(size = 14, face = "bold"), strip.text = element_text(size = 12, face = "bold"))

sound_change_plot_smooth
```

### Sound change by year + gender

```{r}
#create a dummy data frame to store the values from plot_smooth
mod_pred_yobgender_values <- data.frame(plot_smooth(gam_F1_aː, view="participant_year_of_birth", rm.ranef=T)$fv) %>%
   mutate(model1 = "gam_F1_aː1") %>%
  filter(model1 != "gam_F1_aː1")

#load each of the files with for loop
for (i in model_summary_files) {

  #remove .rds from model name and then print the model name
  mod1 <- gsub(".rds", "", i)
  print(mod1)

  #generate a plot smooth of the model giving the participant year of birth*gender interaction smooth
  plot1 <- plot_smooth(get(mod1), view="participant_year_of_birth", plot_all = "Gender", n.grid = 119, rm.ranef = T, main = mod1)

  #extract the values from the plot and reformat for ggplot
  plot1 <- plot1$fv %>%
    mutate(Vowel = substr(mod1, 8, nchar(mod1)),
           Formant = substr(mod1, 5, 6))

  #combine the data to the previous rows, so that all models are in a single data frame
  mod_pred_yobgender_values <<- rbind(mod_pred_yobgender_values, plot1)

}

```

```{r}

mod_pred_yobgender_values <- readRDS("Models/mod_pred_yobgender_values.rds")

#plot the gam smooths predicting F1/F2 per vowel and add the per speaker mean values with F or M as the text
sound_change_plot_smooth_gender <- mod_pred_yobgender_values %>%
  mutate(Formant = factor(Formant, levels = c("F2", "F1"))) %>%
  ggplot(aes(x = participant_year_of_birth, y = fit, colour = Formant, linetype = Gender, label = Gender, fill = Formant)) +
  geom_text(data = gam_intercepts.tmp_long %>%
  mutate(Formant = factor(Formant, levels = c("F2", "F1"))), aes(x = participant_year_of_birth, y = fit + Intercept), size = 1, alpha = 0.5) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = ll, ymax = ul), alpha = 0.2, colour = NA) +
  scale_x_continuous(breaks = c(1900, 1950)) +
  xlab("Year of birth") +
  ylab("Predicted model fit (Lobanov 2.0)") +
  facet_grid(Formant~Vowel) +
  theme_bw() +
  theme(legend.position = "top", axis.text = element_text(size = 12), axis.title = element_text(size = 14, face = "bold"), strip.text = element_text(size = 12, face = "bold")) +
  guides(linetype=guide_legend(override.aes=list(fill=NA)))

sound_change_plot_smooth_gender

ggsave(plot = sound_change_plot_smooth_gender, filename = "Figures/sound_change_gam_gender.png", width = 10, height = 5, dpi = 600)
```

### Visualisation in F1/F2 space

```{r fig.width=4, fig.height=4}
#transform data so there are separate columns for F1 and F2
sound_change_plot_data <- mod_pred_yob_values %>%
  select(participant_year_of_birth, fit, Vowel, Formant) %>%
  pivot_wider(names_from = Formant, values_from = fit) #transform the data to wide format so there are separate F1 and F2 variables

#make data frame to plot starting point, this will give the vowel labels based on the smallest year of birth coordinates
sound_change_labels1 <- sound_change_plot_data %>%
  group_by(Vowel) %>%
  filter(participant_year_of_birth == min(participant_year_of_birth))

#make data frame to plot end point, this will give the arrow at the end of the paths based on the largest year of birth coordinates
sound_change_labels2 <- sound_change_plot_data %>%
  group_by(Vowel) %>%
  top_n(wt = participant_year_of_birth, n = 2)

#plot
sound_change_plot <- sound_change_plot_data %>%
  #set general aesthetics
  ggplot(aes(x = F2, y = F1, colour = Vowel, alpha = participant_year_of_birth, group = Vowel)) +
  #add year of birth change trajectories
  geom_path(size = 0.5, show.legend = FALSE) 
+
  #add end points (this gives the arrows)
  geom_path(data = sound_change_labels2, aes(x = F2, y = F1, colour = Vowel, group = Vowel),
            arrow = arrow(ends = "last", type = "closed", length = unit(0.2, "cm")),
            inherit.aes = FALSE, show.legend = FALSE) +
  #plot the vowel labels
  geom_text(data = sound_change_labels1, aes(x = F2, y = F1, colour = Vowel, group = Vowel, label = Vowel), inherit.aes = FALSE, show.legend = FALSE) +
  #label the axes
  xlab("F2 (normalised)") +
  ylab("F1 (normalised)") +
  #scale the size so the path is not too wide
  scale_size_continuous(range = c(0.2,1)) +
  #reverse the axes to follow conventional vowel plotting
  scale_x_reverse(limits = c(2,-2), position = "top") +
  scale_y_reverse(limits = c(2.3,-2), position = "right") +
  #set the colours
  #scale_color_manual(values = c("#9590FF", "#D89000", "#A3A500", "#39B600", "#00BF7D",
  #                               "#00BFC4", "#00B0F6", "#F8766D", "#E76BF3", "#FF62BC")) +
  scale_color_manual(values = c("#9590FF", "#D89000", "#A3A500")) +  #add a title
  # labs(title = "A) Change over time\n     ") +
  #set the theme
  theme_bw() +
  #make title bold
  theme(plot.title = element_text(face="bold"))

sound_change_plot

ggsave(plot = sound_change_plot, filename = "Figures/sound_change_static.png", width = 5.5, height = 5.5, dpi = 300)

```

## Pillai distance between vowel clouds 

Range 0 to 1. A low Pillai score indicates that the two clouds are merged.

```{r}
# calculate the Pillai score
# measure the degree of overlap
# see also: https://joeystanley.com/blog/pillai-scores-dont-change-after-normalization
# normalisation is not important here

formants_all %>%
  filter(labels == "aː" | labels == "ɑ") %>%
    group_by(id) %>%
    filter(n() >= 20) %>%
  #group_by(Geschlecht, Alter) %>%
  mutate(pillai = pillai(cbind(T1, T2) ~ labels)) %>%
  lm(formula= pillai ~ Alter + Geschlecht) %>%
  summary()

%>%
  arrange(pillai) %>%
#  select(Geschlecht, Alter, pillai) %>%
#  group_by(Geschlecht, Alter) %>%
#  distinct(pillai) %>%
  ggplot(aes(x= Alter, y= pillai, group = Geschlecht)) +
  geom_line(aes(linetype = Geschlecht)) +
  geom_point() +
  ylim(0,1)
```



## Calculate distances for pairs of vowels
```{r}
# distances calculated on the basis of the speaker's median per vowel
distances_df <- formants_speaker_median %>%
  select(labels, id, T1, T2, F1_lob, F2_lob, Alter, Geschlecht) %>%
  pivot_wider(names_from = labels, values_from = c(T1, T2, F1_lob, F2_lob)) %>%
  #na.omit() %>%
  # euclidian distances
  # in Hz
  mutate(euclidean_aː_æːɪ = eucl_dist(T1_aː, T1_æːɪ, T2_aː, T2_æːɪ)) %>%
  mutate(euclidean_æ_æːɪ = eucl_dist(T1_æ, T1_æːɪ, T2_æ, T2_æːɪ)) %>%
  mutate(euclidean_aː_ɑ = eucl_dist(T1_aː, T1_ɑ, T2_aː, T2_ɑ)) %>%
  mutate(euclidean_aː_æ = eucl_dist(T1_aː, T1_æ, T2_aː, T2_æ)) %>%
  mutate(euclidean_ɜɪ_æːɪ = eucl_dist(T1_ɜɪ, T1_æːɪ, T2_ɜɪ, T2_æːɪ))%>%
  mutate(euclidean_ɜɪ_æ = eucl_dist(T1_ɜɪ, T1_æ, T2_ɜɪ, T2_æ)) %>%
  # normalised values
  mutate(euclidean_lob_aː_æːɪ = eucl_dist(F1_lob_aː, F1_lob_æːɪ, F2_lob_aː, F2_lob_æːɪ)) %>%
  mutate(euclidean_lob_æ_æːɪ = eucl_dist(F1_lob_æ, F1_lob_æːɪ, F2_lob_æ, F2_lob_æːɪ)) %>%
  mutate(euclidean_lob_aː_ɑ = eucl_dist(F1_lob_aː, F1_lob_ɑ, F2_lob_aː, F2_lob_ɑ)) %>%
  mutate(euclidean_lob_aː_æ = eucl_dist(F1_lob_aː, F1_lob_æ, F2_lob_aː, F2_lob_æ)) %>%
  mutate(euclidean_lob_ɜɪ_æːɪ = eucl_dist(F1_lob_ɜɪ, F1_lob_æːɪ ,F2_lob_ɜɪ, F2_lob_æːɪ))%>%
  mutate(euclidean_lob_ɜɪ_æ = eucl_dist(F1_lob_ɜɪ ,F1_lob_æ ,F2_lob_ɜɪ,F2_lob_æ)) %>%
  # fronting/backing distances, i.e. relation to F2(i)
  mutate(F2distance_i_æ = F2_lob_i - F2_lob_æ) %>%
  mutate(F2distance_i_æːɪ = F2_lob_i - F2_lob_æːɪ) %>%
  mutate(F2distance_i_æːʊ = F2_lob_i - F2_lob_æːʊ) %>%
  mutate(F2distance_i_ɑ = F2_lob_i - F2_lob_ɑ) %>%
  mutate(F2distance_i_ɑɪ = F2_lob_i - F2_lob_ɑɪ) %>%
  mutate(F2distance_i_ɑʊ = F2_lob_i - F2_lob_ɑʊ) %>%
  mutate(F2distance_i_aː = F2_lob_i - F2_lob_aː) %>%
  # raising / lowering distances, i.e. relation to F1(i)
  mutate(F1distance_i_aː = F1_lob_aː - F1_lob_i) %>%
  mutate(F1distance_i_ɑ = F1_lob_ɑ - F1_lob_i) %>%
  mutate(F1distance_i_æ = F1_lob_æ - F1_lob_i) %>%
  mutate(F1distance_i_ɑʊ = F1_lob_ɑʊ - F1_lob_i) 

ggplot(distances_df, aes(x= Alter, y = F1distance_i_æ)) +
  geom_boxplot() +
  facet_wrap(~ Geschlecht)
  # for regression
  #geom_point() +
  # regression line
  #stat_summary(geom = "line", fun = median, group = 1)

# regression analysis
regression <- lm(F1distance_i_æ ~ Alter + Geschlecht,  data = distances_df)
summary(regression)
```


##  Duration by age
```{r}
duration_df <- formants_speaker_median %>%
  select(labels, id, duration, Alter, Geschlecht) %>%
  pivot_wider(names_from = labels, values_from = duration, names_prefix = "dur_") %>%
  mutate(dist_dur_æːɪ_æːʊ = (dur_æːɪ - dur_æːʊ)) %>%
  mutate(dist_dur_æːɪ_ɑɪ = (dur_æːɪ - dur_ɑɪ)) %>%
  mutate(dur_dist_æːɪ_ɜɪ = (dur_æːɪ - dur_ɜɪ))

regression <- lm(dist_dur_æːɪ_ɑɪ ~ Alter,  data = duration_df)
summary(regression)

ggplot(duration_df, aes(x = Alter, y = dist_dur_æːɪ_ɑɪ)) + 
  geom_boxplot()

```
## Regression for duration

```{r}
library(lme4)

lm <- lm(duration ~ labels*Alter, data = vowel_midpoints)
summary(lm)
```

# Diphthongs

# Data wrangling for the diphthongs
```{r}
formants_onset_diphthongs_0.2 <- get_formants(data = td_vowels_norm, 
                                          time = 0.2,
                                          segments = "æːɪ|æːʊ|ɑɪ|ɑʊ|ɜɪ|əʊ|iə|uə")
formants_onset_diphthongs_0.35 <- get_formants(data = td_vowels_norm, 
                                          time = 0.35,
                                          segments = "æːɪ|æːʊ|ɑɪ|ɑʊ|ɜɪ|əʊ|iə|uə")
formants_onset_diphthongs_0.5 <- get_formants(data = td_vowels_norm, 
                                          time = 0.5,
                                          segments = "æːɪ|æːʊ|ɑɪ|ɑʊ|ɜɪ|əʊ|iə|uə")
formants_onset_diphthongs_0.65 <- get_formants(data = td_vowels_norm, 
                                          time = 0.65,
                                          segments = "æːɪ|æːʊ|ɑɪ|ɑʊ|ɜɪ|əʊ|iə|uə")
formants_onset_diphthongs_0.8 <- get_formants(data = td_vowels_norm, 
                                          time = 0.8,
                                          segments = "æːɪ|æːʊ|ɑɪ|ɑʊ|ɜɪ|əʊ|iə|uə")

formants_diphthongs <- bind_rows(formants_onset_diphthongs_0.2, formants_onset_diphthongs_0.35, formants_onset_diphthongs_0.5, formants_onset_diphthongs_0.65, formants_onset_diphthongs_0.8) %>%
    # create new vowel labels (old label + times_norm) to treat these as separate vowels in Lobanov normalisation 
    dplyr::mutate(labels = paste0(labels, times_norm))

# df with all segments for analysis
# needs the formants_mid_monophthongs df to use the full vowel system for normalisation
# all diphthong measurements have their own label, p.ex. :ɑɪ0.35
formants_dipthongs <- bind_rows(formants_mid_monophthongs, formants_diphthongs) %>%
    # create id for speaker
    mutate(id = word(bundle, 3, sep ="_")) %>%
    # join Sozialdaten
    inner_join(sozialdaten) %>%
    # remove outliers
    #group_by(Alter, Geschlecht, labels) %>%
    #filter(!find_outliers(T1, T2, keep = 0.95)) %>%
    #ungroup() %>%
    # normalise with Lobanov per id, i.e. on the basis of the individual speaker
    group_by(id) %>%
    mutate(F1_lob = normLobanov(T1), F2_lob = normLobanov(T2), F3_lob = normLobanov(T3)) %>%
    ungroup() %>%
    # keep only speakers with more than 40 (?) segments  
    group_by(id) %>%
    filter(n() >= 30) %>%
    ungroup()

# split the 5 measurement points per diphthong again, additional column for 'step' (= times_norm)
formant_tracks_diphthongs <- formants_diphthongs %>%
  # all 5 measurements have a '0' in their name
  filter(str_detect(labels, "0")) %>%
  separate(labels, sep= "0", into = c("labels", "step"))
```


```{r}
data <- formant_tracks_diphthongs %>%
  filter(labels %in% c("ɑɪ", "æːɪ", "ɜɪ", "əʊ", "ɑʊ", "æːʊ", "iə" , "uə")) %>%
  # create id for speaker
  mutate(id = word(bundle, 3, sep ="_")) %>%
  # join Sozialdaten
  inner_join(sozialdaten) %>%
  #filter(Geschlecht == "Weiblech") %>%
  group_by(Alter, Geschlecht, labels, times_norm) %>%
  summarise(F1 = median(T1), F2 = median(T2), F1_lob = median(F1_lob), F2_lob = median(F2_lob)) %>%
  arrange(by_group = TRUE) %>%
  ungroup()

ggplot(data, aes(x = F2_lob, y = F1_lob, col = labels)) +
  geom_path(lineend = "round", size = 1, arrow = arrow(ends = "last", type = "open", length = unit(0.1, "inches"))) +
  scale_x_reverse() +
  scale_y_reverse() +
  facet_wrap(. ~ Geschlecht + Alter, nrow = 2)

```



## Old stuff ##

## Regression

```{r}
library(lme4)

euclid_df <- euclid_df %>%
  mutate(id = as.integer(id)) %>%
  drop_na(id) %>%
  filter(euclidean <= 2 | Dialektgebiet =="") 
  # when working with Alter als ordered factor
  # https://stackoverflow.com/questions/25735636/interpretation-of-ordered-and-non-ordered-factors-vs-numerical-predictors-in-m
  # mutate(Alter = factor(Alter, levels = c("≤ 24", "25 bis 34", "35 bis 44", "45 bis 54", "55 bis 64", "65+"), ordered = TRUE))

#rio::export(euclid_df, "euclid_df.csv")

lm <- lm(euclidean ~ Alter * Geschlecht, data = euclid_df)
summary(lm)
```

If only the Intercept is significant, this means that the predictors chosen have no influence on the dependent variable.


## Random forest

```{r}
set.seed(159)

euclid_split <- initial_split(euclid_df, 
                                prop = 3/4)
# extract training and testing sets
euclid_train <- training(euclid_split)
euclid_test <- testing(euclid_split)

# create CV object from training data
euclid_cv <- vfold_cv(euclid_train)

# define the recipe
euclid_recipe <- 
  # which consists of the formula (outcome ~ predictors)
  recipe(euclidean ~ Alter + Geschlecht + Dialektgebiet, 
         data = euclid_df) %>%
  # and some pre-processing steps
  step_normalize(all_numeric()) %>%
  step_impute_knn(all_predictors())

euclid_train_preprocessed <- euclid_recipe %>%
  # apply the recipe to the training data
  prep(euclid_train) %>%
  # extract the pre-processed training dataset
  juice()

# baseline model
# you want to train a very simple baseline model and see if a more complex model
# improves prediction accuracy.
# you need to do this on the preprocessed training data, and then evaluate it
# on the testing data

baseline_lm <- lm(euclidean ~ Alter + Geschlecht + Dialektgebiet, 
         data = euclid_train_preprocessed)

# use the model to predict on the testing set
predictions <- predict.lm(baseline_lm, euclid_test) %>%
  bind_cols(preds = ., euclid_test) 
#predict.lm creates a vector of predictions. I add this as a column to the data set

# now compute rmse on the baseline model
yardstick::rmse(predictions, truth = euclidean, estimate = preds)

# now you can compare to the random forest below

rf_model <- 
  # specify that the model is a random forest
  rand_forest() %>%
  # specify that the `mtry` parameter needs to be tuned
  set_args(mtry = tune(), tree=tune()) %>%
  # select the engine/package that underlies the model
  set_engine("ranger", importance = "impurity") %>%
  # choose either the continuous regression or binary classification mode
  set_mode("regression") 

# set the workflow
rf_workflow <- workflow() %>%
  # add the recipe
  add_recipe(euclid_recipe) %>%
  # add the model
  add_model(rf_model)

# specify which values eant to try
rf_grid <- expand.grid(mtry = c(3, 4, 5))
# extract results
rf_tune_results <- rf_workflow %>%
  tune_grid(resamples = euclid_cv, #CV object
            grid = rf_grid, # grid of values to try
            #metrics = metric_set(accuracy, roc_auc) # this is wrong, this is used for classification, not regression
            metrics = metric_set(rmse) #
            )

# explainability
explain_rf <- DALEX::explain(model = rf_tune_results,  
                          data = euclid_df,
                             y = euclid_df$euclidean, 
                         label = "Random Forest")



# importance
final_model <- fit(rf_workflow, euclid_df)

ranger_obj <- pull_workflow_fit(final_model)$fit
ranger_obj
ranger_obj$variable.importance

```

